---
title: 'Appendix - Selection against instability: stable subgraphs are most frequent
  in empirical food webs'
author: "Jonathan J. Borrelli"
date: "Wednesday, July 23, 2014"
output:
  pdf_document:
    toc: yes
    number_sections: yes
    highlight: tango
  html_document:
    number_sections: yes
    toc: yes
---

```{r, echo = F, message = F}
require(knitr)
opts_chunk$set(message = F, comment = NA, tidy = T)

```

--------------------------------------------------  


Borrelli, J. J. 2014. Selection against instability: stable subgraphs are most frequent in empirical food webs. -Oikos 000: 000-000 
  
[The code provided here can also be obtained from GitHub by clicking this link](https://github.com/borre_000/Subgraph-Stability)

# Definitions  

## Subgraph Library  

The following code defines the sign matrix for each of the thirteen possible three-node subgraphs. Here, the "s" in the object name indicates that only single links are used, while a "d" indicates the presence of double links.   

```{r library}
s1<-matrix(c(0,1,0,-1,0,1,0,-1,0),nrow=3,ncol=3)
s2<-matrix(c(0,1,1,-1,0,1,-1,-1,0),nrow=3,ncol=3)
s3<-matrix(c(0,1,-1,-1,0,1,1,-1,0),nrow=3,ncol=3)
s4<-matrix(c(0,1,1,-1,0,0,-1,0,0),nrow=3,ncol=3)
s5<-matrix(c(0,0,1,0,0,1,-1,-1,0),nrow=3,ncol=3)

d1<-matrix(c(0,1,1,-1,0,1,-1,1,0),nrow=3,ncol=3)
d2<-matrix(c(0,1,1,1,0,1,-1,-1,0),nrow=3,ncol=3)
d3<-matrix(c(0,1,-1,1,0,0,1,0,0),nrow=3,ncol=3)
d4<-matrix(c(0,1,1,-1,0,0,1,0,0),nrow=3,ncol=3)
d5<-matrix(c(0,1,1,-1,0,1,1,-1,0),nrow=3,ncol=3)
d6<-matrix(c(0,1,1,1,0,1,1,1,0),nrow=3,ncol=3)
d7<-matrix(c(0,1,1,1,0,1,1,-1,0),nrow=3,ncol=3)
d8<-matrix(c(0,1,1,1,0,0,1,0,0),nrow=3,ncol=3)

mot.lst <- list(s1, s2, s3, s4, s5, d1, d2, d3, d4, d5, d6, d7, d8)
names(mot.lst) <- c("s1", "s2", "s3", "s4", "s5", "d1", "d2", "d3", "d4", "d5", "d6", "d7", "d8")

```

## Define required functions

### Functions for counting motifs

The `motif_counter` function takes in a list of graph objects and applies `triad.census` to each. It returns a data frame of the frequency of each connected three-node digraph. 

```{r counter}
motif_counter <- function(graph.lists){
  require(igraph)
  
  if(!is.list(graph.lists)){
    stop("The input should be a list of graph objects")
  }
  
  triad.count <- lapply(graph.lists, triad.census)
  triad.matrix <- matrix(unlist(triad.count), nrow = length(graph.lists), ncol = 16, byrow = T)
  colnames(triad.matrix) <- c("empty", "single", "mutual", "s5", "s4", "s1", "d4",
                              "d3", "s2", "s3","d8", "d2", "d1", "d5", "d7", "d6")
  
  triad.df <- as.data.frame(triad.matrix)
  
  motif.data.frame <- data.frame(s1 = triad.df$s1, s2 = triad.df$s2, s3 = triad.df$s3, s4 = triad.df$s4, 
                                 s5 = triad.df$s5, d1 = triad.df$d1, d2 = triad.df$d2, d3 = triad.df$d3, d4 = triad.df$d4,
                                 d5 = triad.df$d5, d6 = triad.df$d6, d7 = triad.df$d7, d8 = triad.df$d8)
  
  rownames(motif.data.frame) <- names(graph.lists)
  return(motif.data.frame)
}
```

The Curveball algorithm is available as supplemental information as part of the original publication. *Note if you want to use this code please cite the paper in which it was introduced:*  

> Strona, G. et al. 2014. A fast and unbiased procedure to randomize ecological binary matrices with fixed row and column totals. -Nat. Comm. 5: 4114. [doi: 10.1038/ncomms5114](http://www.nature.com/ncomms/2014/140611/ncomms5114/full/ncomms5114.html)   

Their function takes a matrix and makes a swap (process descirbed in their paper) and returns the new matrix.   

```{r}
curve_ball<-function(m){
  RC=dim(m)
  R=RC[1]
  C=RC[2]
  hp=list()
  for (row in 1:dim(m)[1]) {hp[[row]]=(which(m[row,]==1))}
  l_hp=length(hp)
  for (rep in 1:5*l_hp){
    AB=sample(1:l_hp,2)
    a=hp[[AB[1]]]
    b=hp[[AB[2]]]
    ab=intersect(a,b)
    l_ab=length(ab)
    l_a=length(a)
    l_b=length(b)
    if ((l_ab %in% c(l_a,l_b))==F){
      tot=setdiff(c(a,b),ab)
      l_tot=length(tot)
      tot=sample(tot, l_tot, replace = FALSE, prob = NULL)
      L=l_a-l_ab
      hp[[AB[1]]] = c(ab,tot[1:L])
      hp[[AB[2]]] = c(ab,tot[(L+1):l_tot])}
    
  }
  rm=matrix(0,R,C)
  for (row in 1:R){rm[row,hp[[row]]]=1}
  rm
}
```

The `curving` function is used to iteratively apply the Curveball algorithm to a single matrix. This function takes an adjacency matrix and number of iterations as inputs and returns a dataframe of motif frequencies.   
 
```{r}
curving <- function(adjmat, n){
  mot <- motif_counter(list(graph.adjacency(adjmat)))
  newmat <- adjmat
  
  for(i in 1:n){
    newmat <- curve_ball(newmat)
    m <- motif_counter(list(graph.adjacency(newmat)))
    mot <- rbind(mot, m)
  }
  return(mot[-1,])
}
```

I added the additional constraints of maintaining the number of single, double, and self links in each matrix to the original curveball algorithm in the function `dblcan.curve`. This function takes a matrix and desired number of iterations as inputs and returns a dataframe of motif frequencies.   

```{r}
library(plyr)

nd <- function(gl) nrow(gl) - nrow(unique(aaply(gl, 1, sort)))
# determines the number of double links

dblcan.curve <- function(mat, iter){
  mot <- motif_counter(list(graph.adjacency(mat)))
  
  el <- get.edgelist(graph.adjacency(mat))
  Ne <- nrow(el)
  dbl <- nd(el)
  can <- sum(diag(mat))

  for(i in 1:iter){
    ed = TRUE
    dub = TRUE
    ca = TRUE
    while(ed || dub || ca){
      mat2 <- curve_ball(mat)
      
      el2 <- get.edgelist(graph.adjacency(mat2))
      el3 <- unique(el2)
      Ne2 <- nrow(el3)
      dbl2 <- nd(el3)
      can2 <- sum(diag(mat2))
      
      ed <- Ne != Ne2
      dub <- dbl != dbl2
      ca <- can != can2
      
    }
    mat <- mat2
    
    mot <- rbind(mot, motif_counter(list(graph.adjacency(mat))))
  }
  return(M = mot[-1,])
}
```


### Functions for determining quasi sign-stability
There are two main functions for determining quasi sign-stability, and a third that wraps them together to generate the desired number of iterations. 

The function `ran.unif` takes an input of a signed matrix. It will then check each cell to see if there is a 1 or -1. Each 1 will be replaced by a value drawn from the random uniform distribution between 0 and 10, while each -1 is replaced by a value from the random uniform distribution between -1 and 0. The `ran.unif` function also assigns values to the diagonal from a random uniform distribition between -1 and 0. The resulting randomly sample matrix is returned.  

```{r}
ran.unif <- function(motmat){
  newmat <- apply(motmat, c(1,2), function(x){
    if(x==1){runif(1, 0, 10)}else if(x==-1){runif(1, -1, 0)} else{0}
  })
  diag(newmat) <- runif(3, -1, 0)
  return(newmat)
}
```

Given the input matrix `maxRE` will compute the eigenvalues and return the largest real part. 

```{r}
maxRE <- function(rmat){
  lam.max <- max(Re(eigen(rmat)$values))
  return(lam.max)
}

```

The above two functions are combined in `eig.analysis`. Given the number of desired sampling iterations, `n`, and a list of sign matrices to analyze, `matrices`, the `eig.analysis` function will return an  `n` by `length(matrices)` matrix of eigenvalues. Specifically it is returning the $max(Re(\lambda))$ for each sampled matrix. From this matrix quasi sign-stability can be calculated as the proportion of values in each column that are negative.  

```{r}
eig.analysis <- function(n, matrices){
  cols <- length(matrices)
  rows <- n
  eigenMATRIX <- matrix(0, nrow = rows, ncol = cols)
  for(i in 1:n){
    ranmat <- lapply(matrices, ran.unif)
    eigs <- sapply(ranmat, maxRE)
    eigenMATRIX[i,] <- eigs
  }
  return(eigenMATRIX)
}

```

# Analysis  

Load required packages
```{r}
library(igraph)
library(ggplot2)
library(reshape2)
library(parallel)
library(doSNOW)
```


## Determining motif frequency

Load in web data from GitHub. [Click here to download the .Rdata file](https://github.com/borre_000/Subgraph-Stability/blob/master/webGRAPHS.Rdata?raw=true). This file is a list of igraph graph objects for each of the 50 webs used in the analysis. Once you have downloaded the file into your working directory:

```{r wd, echo = F}
setwd("C:/Users/borre_000/Desktop/Github/Subgraph-Stability/")

```

```{r getDATA}
load(paste(getwd(), "webGRAPHS.Rdata", sep = "/"))
```

The frequencies of each of the different subgraphs can now be determined easily with `motif_counter`. 

```{r}
motfreq <- motif_counter(web.graphs)
kable(motfreq, format = "pandoc")
```

The following code runs the null model analysis for the 50 food webs. First, each of the fifty webs are converted into binary adjacency matrices (`web.matrices`).    
```{r eval = F}
web.matrices <- lapply(web.graphs, get.adjacency, sparse = F)
```

The first null model, the Curveball algorithm can be applied to all 50 webs in parallel. This code starts by registering a cluster utilizing 1 less than the number of cores on the computer (for this paper it was 7 cores). It then creates set of 30000 matrices and returns a list (length equal to the number of webs, 50 in this case) of dataframes of motif frequences.  

```{r, eval = F}
cl <- makeCluster(detectCores()-1)
clusterExport(cl, c("web.adj", "motif_counter", "curve_ball", "curving"))
registerDoSNOW(cl)
randos <- parLapply(cl, web.adj, curving, n = 30000)
stopCluster(cl)
```


Once subgraph counts have been obtained, the mean and standard deviation for each subgraph are computed. Z-scores are then computed as described in the methods section:  
$$
z_i = \frac{x_i - \overline{x}}{\sigma}
$$
The normalized profile was then computed (as desribed in the methods):  
$$
n_i = \frac{z_i}{\sqrt{\sum{z_j^2}}}
$$

```{r eval = F}
means <- t(sapply(randos, colMeans))
stdevs <- t(sapply(randos, function(x){apply(x, 2, sd)}))

motfreq <- motif_counter(web.graphs)

zscore <- (motfreq - means)/stdevs

# Normalized z-scores
zscore.norm <- t(apply(zscore, 1, function(x){x/sqrt(sum(x^2, na.rm = T))}))
```

Below is the code used to run the Curveball algorithm with the additional constraints described above. *Note this can take a very long time to run*
```{r eval = F}
# Use the filepath.sink variable to set the path for
# storing the sink information 
filepath.sink <- my.sink.path
# Use the filepath.data variable to set the location for 
# storing the dataframes of motif frequencies for each web
filepath.data <- my.data.path

cl <- makeCluster(detectCores()-1)
clusterExport(cl, c("web.adj", "motif_counter", "curve_ball", "nd", "aaply", "filepath"))
registerDoSNOW(cl)
system.time(
randos.t3 <- foreach(i = 1:length(web.adj)) %dopar% {
  sink(file = paste(filepath.sink, names(web.adj[i]), ".txt", collapse = ""))
  motfreq <- dblcan.curve(web.adj[[i]], iter = 30000)
  print(motfreq)
  sink()
  write.csv(motfreq, file = paste(filepath.data, names(web.adj[i]), ".csv", collapse = ""))
  return(motfreq)
}
)
stopCluster(cl)
```

The normalized z-score profile can be calculated the same as above: 
```{r eval = F}
means.t <- t(sapply(randos.t3, colMeans))
stdevs.t <- t(sapply(randos.t3, function(x){apply(x, 2, sd)}))

motfreq <- motif_counter(web.graphs)

zscore.t <- (motfreq - means.t)/stdevs.t

# Normalized z-scores
zscore.N <- t(apply(zscore.t, 1, function(x){x/sqrt(sum(x^2, na.rm = T))}))
```


**Figure 1** is then a boxplot of the above normalized z-scores, reordered according to decreasing quasi sign-stability (see below). When using the modified Curveball algorithm, those webs that have no double links produce `NaN` when computing the z-score for those motifs. These are ignored in **Figure 1**, and the z-scores presented are with the `NaN`s  removed.   

## Determining Quasi Sign-Stability  

The first step to get quasi sign stability is to get the largest eigenvalues from a series of randomly parameterized sign matrices. In the following code I generate 10000 random parameterizations for each of the 13 subgraphs's sign matrices (`mot.lst`). The `eig.analysis` function will return a matrix where each column is a different subgraph and each row is the largest eigenvalue of a particular randomization. 

```{r eigs, cache = T}
set.seed(5)

n <- 10000
mot.stab<- eig.analysis(n, mot.lst)
colnames(mot.stab) <- names(mot.lst)

```

From that matrix, quasi sign-stability is calculated as the proportion of rows with a negative value. In other words, how many random parameterizations of the sign matrix were locally stable? 

```{r qss}
mot.qss <- apply(mot.stab, 2, function(x){sum(x<0)/n})
sorted <- sort(mot.qss, decreasing = T)
sorted
```

### Robustness to assumption of double link positives
I assumed that if there was a double link in the subgraph, that both the effect of the predator on the prey and the effect of the prey on the predator were positive, a (+/+) rather than a (+/-). Here I repeat the analysis described above, but instead assuming that the relative effects correspond to a (-/-). 

```{r, cache = T}
mot.lst2 <- lapply(mot.lst, function(x){
  for(i in 1:nrow(x)){
    for(j in 1:ncol(x)){
      if(x[i,j] == 1 && x[j,i] == 1){x[i,j] <- x[i,j]*-1; x[j,i] <- x[j,i]*-1}else{next}
    }
  }
  return(x)
})

set.seed(15)

n <- 10000
mot.stab2 <- eig.analysis(n, mot.lst2)
colnames(mot.stab2) <- names(mot.lst2)

mot.qss2 <- apply(mot.stab2, 2, function(x){sum(x<0)/n})
sorted2 <- sort(mot.qss2, decreasing = T)
sorted2
```

### Robustness to assumption of matrix fill distributions
I also tested how quasi sign-stability of the different subgraphs changed when varying the assumption of the relative impact of the predator on its prey and the prey on its predator. 

I tested 7 additional distributions (all uniform) with different magnitudes. 
```{r warning=FALSE}
params.u <- data.frame(pred1 = c(0, 0, 0, 0, 0, 0, 0, 0), pred2 = c(10, 10, 10, 10, 10, 5, 3, 1), 
                       prey1 = c(-10, -5, -1, -.1, -.01, -1, -1, -1), prey2 = c(0, 0, 0, 0, 0, 0, 0, 0))
parvals <- factor(paste(params.u[,2], params.u[,3], sep = "/"), 
                  levels = c("1/-1", "3/-1", "5/-1", "10/-1", "10/-0.01", "10/-0.1", "10/-1", "10/-5", "10/-10"))
```

Below is a function to take in the different parameters for the distributions and fill a matrix accordingly. 
```{r}
eigen_unif <- function(m, iter, params, self = -1){
  # For when I want to use uniform distribution
  # Params is dataframe of min and max for relative impact of prey on pred 
  ## and pred on prey
  ev <- c()
  for(i in 1:iter){
    m1 <- apply(m, c(1,2), function(x){
      if(x==1){runif(1, params$pred1, params$pred2)}else if(x==-1){runif(1, params$prey1, params$prey2)} else{0}
    })
    diag(m1) <- self
    ev[i] <- max(Re(eigen(m1)$values)) 
  }
  return(ev)
}
```

The following code loops through the dataframe of distributions and computes quasi sign-stability
```{r cache = T}
test <- matrix(nrow = nrow(params.u), ncol = length(mot.lst))
for(i in 1:nrow(params.u)){
  eigen.test <- lapply(mot.lst, eigen_unif, iter = 10000, 
                       params = params.u[i,], self = runif(3, -1, 0))
  qss.test <- lapply(eigen.test, function(x){
    sum(x < 0) / length(x)
  })
  test[i,] <- unlist(qss.test)
}
```


### Figure A1

```{r warning=F}
colnames(test) <- c("s1", "s2", "s3", "s4", "s5", "d1", "d2", "d3", "d4", "d5", "d6", "d7", "d8")
test <- data.frame(test, parvals)
dat1 <- melt(test[,c(names(sorted),"parvals")])

ggplot(dat1, aes(x = variable, y = value)) + geom_point() + facet_wrap(~parvals, ncol = 4) + theme_bw() + xlab("Subgraph") + ylab("Quasi Sign-Stability")
```
**Figure A1: Robustness of quasi sign-stability to different distributions used to fill the random Jacobian matrix. Facet labels represent the maximum impact of the prey on the predator population (positive number) and maximum impact of the predator on the prey population (negative number)**


# Code for the figures

## Figure 1
```{r fig.height = 1, fig.width = 7}
par(mfrow = c(1, 13), mar = c(.2, .2, .2, .2))
for(i in 1:13){
  plot.igraph(graph.adjacency(mot.lst[[which(names(mot.lst) == names(sorted[i]))]]), layout = layout.circle, edge.arrow.size = .5, 
              vertex.size = 30, vertex.color = "black", vertex.label = NA, frame = T, edge.width = 2, edge.color = "darkslategray4")
  text(-.25,0,names(sorted[i]), cex = 1.5)
}
```


## Figure 2
```{r echo = F}
z.both <- read.csv("C:/Users/borre_000/Desktop/GitHub/Subgraph-Stability/zboth.csv", row.names = 1)
```


```{r eval = F}
z1 <- cbind(Model = factor("Curveball"), melt(zscore.norm[,names(sorted)]))
z2 <- cbind(Model = factor("Double"), melt(zscore.N[,names(sorted)]))

z.both <- rbind(z1, z2)
ggplot(z.both, aes(x = Var2, y = value, fill = Model)) + geom_boxplot() + xlab("Subgraph") + ylab("Normalized Profile") + theme_bw() 
```

```{r echo = F}
z.both$Var2 <- factor(z.both$Var2, levels = names(sorted))
ggplot(z.both, aes(x = Var2, y = value, fill = Model)) + geom_boxplot() + xlab("Subgraph") + ylab("Normalized Profile") + theme_bw()
```

## Figure 3
```{r}
sort.df <- melt(sorted)

qssplot <- ggplot(sort.df, aes(x = 1:13, y = value)) + geom_point(shape = 19, size = 3) + theme_bw()
qssplot + xlab("Subgraph") + ylab("Quasi Sign-Stability") + scale_x_discrete(limits=names(sorted))
```



